\documentclass{template}
\title{Summary - Probability Theory}

\newcommand{\Poisson}{\text{Poisson}}
\newcommand{\Binomial}{\text{Bin}}
\newcommand{\ChiSquare}{\chi^2}
\newcommand{\Exponential}{\text{Exp}}
\newcommand{\Cov}{\text{Cov}}

\begin{document}
\maketitle

\section{Probability}

\subsection{Concepts}

The probability of an certain event $A$ happened is $1$, but an
event with probability $1$ does not necessarily mean that it will happen.
The probability of an impossible event is $0$, but an event with probability $0$ does
not necessarily mean that it will not happen.
\begin{example}
	Let $A$ be the event that a randomly chosen real number from the interval
	$[0, 1]$ is irrational. Then $P(A) = 1$, but it is possible to choose a rational
	number from the interval $[0, 1]$.
	And let $B=\overline{A}$ be the event that a randomly chosen real number from the interval,
	then $P(B) = 0$, but it is possible to choose a rational number from the interval $[0, 1]$.
\end{example}

The events set property can imply the Probability relationships, but
the inverse is not true. We can not conclude events set property from the Probability
relationships.

\begin{example}
	Knowing that events $A$ and $B$ satisfy $P(B|A)=1$, we can not conclude that
	$A \subseteq B$. 

	Let $\Omega=[0,1]$ and $A=[0,1], B=[0,1]\backslash \{.5\}$. Then we have $P(B|A)=1$, but
	$A \not\subseteq B$.
\end{example}

\begin{theorem}
0 Probability and 1 Probability events are independent of any other events.
\begin{proof}
	Let $A$ be an event with $P(A)=0$, then for any event $B$,
	\[
		P(A \cap B) \leq P(A) = 0 \implies P(A  B) = 0 = P(A)P(B)
	\]
	Let $A$ be an event with $P(A)=1$, then for any event $B$,
	\[
		B = (A \cap B) \cup (\overline{A} \cap B) \implies P(B) = P(A \cap B) + P(\overline{A} \cap B)
	\]
	since $P(\overline{A}) = 0$, we have
	\[
		P(A \cap B) = P(B) = P(A)P(B)
	\] 
\end{proof}
\end{theorem}

\begin{theorem}
	If $A$ and $B$ are independent and exclusive events, then $P(A)=0$ or $P(B)=0$.
	\begin{proof}
		\[
			P(AB)=P(A)P(B)=0 \implies P(A)=0 \text{ or } P(B)=0
		\] 
	\end{proof}
\end{theorem}
\begin{remark}
	Inversely, if $P(A)>0$ and $P(B)>0$, then $A$ and $B$ can not be both independent and exclusive.
\end{remark}


\subsection{Equations}
\begin{theorem}
	\[
		A B = \overline{A} \cap \overline{B} \iff B = \overline{A}
	\] 
	\begin{proof}
		\[
			\begin{aligned}
				A B = \overline{A} \cap \overline{B} 
				&\iff AB \cup A \overline{B} = (\overline{A} \cap \overline{B}) \cup A \overline{B} \\
				&\iff A(B \cup \overline{B}) = (\overline{A} \cup A) \cap \overline{B} \\
				&\iff A = \overline{B}
			\end{aligned}
		\]
	\end{proof}
\end{theorem}

\subsection{Inequalities}

\[
P(A) \geq P(AB) \quad P(B) \geq P(AB)
\] 
or equivalently,
\[
	P(AB) \leq \min\{P(A), P(B)\}
\] 
\[
	P(AB) \leq \frac{P(A) + P(B)}{2}
\] 
\[ P(AB) = P(A) + P(B) - P(A \cup B) \geq P(A) + P(B) - 1 \] 

\section{Distributions}

\subsection{Distribution Functions}

Given a random variable \( X \) with PDF \( f(x) \), and CDF \( F(x) \) .
\[
	P(X<x) = F(x^{-}) \quad P(X=x) = F(x) - F(x^{-})
\] 

\subsubsection{Transformation of Distribution Functions}

\begin{theorem}
The linear combination of CDFs(or PDFs)
\[ \sum_{i=1}^{n} a_i F_i(x) \]
is also a CDF(or PDFs) if and only if 
\[
	\sum_{i=1}^{n} a_i = 1 \quad a_i \geq 0
\]
\end{theorem}

\begin{example}
	Find $E[X]$ if the PDF of $X$ is given by
	 \[
		 F(x) = 0.3\Phi(x) + 0.7\Phi\left(\frac{x-1}{2}\right)
	\] 
	\begin{answer}
		\[
			E[X] = 0.3E[Z] + 0.7E[2Z+1] = 0.3\cdot 0 + 0.7\cdot (2\cdot 0 + 1) = 0.7
		\] 
	\end{answer}
\end{example}

\begin{theorem}
The multiplication of CDFs $F_1F_2$ is also a CDF, while the multiplication of PDFs
 $f_1f_2$ is not necessarily a PDF.
\end{theorem}

\begin{theorem}
	Given a CDF \( F(x) ,  k \in \mathbb{Z}^{+}\) then
	\[
		F(ax^{2k-1}+b) \text{ is a CDF} \iff a>0
	\] 
	while
	\[
		F(ax^{2k}+b) \text{ never be a CDF for any } a
	\] 
	\begin{proof}
		For the first case, when a is positive, the new function keeps the monotonicity
		and right continuity of the original CDF, and the limits at infinity are preserved.
		So the new function is also a CDF.
		But when a is negative, the new function reverses the monotonicity of the original CDF,
		violating the property of CDF.
		For the second case, we have
		\[
			F(ax^{2k}+b) \to 1 \text{ as } x \to \infty, \quad F(ax^{2k}+b) \to 1 \text{ as } x \to -\infty
		\] 
		violating the property of CDF.
	\end{proof}
\end{theorem}
\begin{theorem}
	Given a PDF \( f(x) \) then
	\[
		f(ax+b) \text{ is also a PDF} \iff |a|=1
	\] 
\end{theorem}

\begin{theorem}
	Given a CDF \( F(x) \) then
	\[
		1-F(-x) \text{ is also a CDF} \iff X \text{ is continuous rv}
	\] 
\end{theorem}

Given two pdf \( f_1 \) and \( f_2 \), and their corresponding CDFs \( F_1 \) and \( F_2 \), then
\begin{theorem}
	\[
		f_1F_2 + f_2F_1 \text{ is also a PDF}
	\] 
	\begin{proof}
		Here is a not rigorous proof using probability method.
		Let  $X_1 \sim f_1, X_2 \sim f_2$ be two independent random variables, consider a new random variable
		\[
		Y = \max(X_1, X_2)
		\] 
		Calculate the CDF of $Y$
		\[
		F_Y(y) = P(Y \leq y) = P(X_1 \leq y, X_2 \leq y) = P(X_1 \leq y)P(X_2 \leq y) = F_1(y)F_2(y)
		\] 
		Take the derivative to get the PDF of $Y$
		\[
		f_Y(y) = \frac{\dd}{\dd y} F_Y(y) = f_1(y)F_2(y) + f_2(y)F_1(y)
		\]
		hence, $f_1F_2 + f_2F_1$ is also a PDF.
	\end{proof}
\end{theorem}
\begin{theorem}
	\[
		2f_1F_1 \text{ and } 2f_2F_2 \text{ are also PDFs}
	\] 
	
\end{theorem}


\subsection{Concepts}

\subsubsection{Percentile}
Let $p$ be a number between 0 and 1, the $100p$-th percentile of a rv $X$
denoted by $\eta(p)$ is defined by
\[
	P(X \leq \eta(p)) = p
\] 
or equivalently,
\[
	p = F(\eta(p)) = \int_{-\infty}^{\eta(p)} f(x) dx
\] 
\subsubsection{Critical Value of Standard Normal Distribution}
$z_\alpha$ is defined by $P(Z \geq z_\alpha) = \alpha$

In other words, $z_\alpha$ is the $100(1-\alpha)$-th percentile of the standard normal distribution.

\begin{center}
\begin{tikzpicture}
	\begin{axis}[
		xmin=-4, xmax=4,
		ymin=0, ymax=0.5,
		axis lines=middle,
		enlargelimits,
		ticks=none,
		width=10cm,
		height=6cm,
		domain=-4:4,
		samples=100,
		ylabel={$f_Z(z)$},
		xlabel={$z$},
	]
	\addplot[blue, thick] {1/sqrt(2*pi) * exp(-x^2/2)};
	\addplot[fill=blue!20, opacity=0.5] {1/sqrt(2*pi) * exp(-x^2/2)} \closedcycle;
	% fill the area to the right of z_alpha
	\addplot[
			fill=cyan!20, opacity=0.8,
			draw=none,
			domain=1.96:4,
			samples=200
	] {1/sqrt(2*pi) * exp(-x^2/2)} \closedcycle;

	\draw[blue,thick] (axis cs: 1.96,0) -- (axis cs: 1.96,{1/sqrt(2*pi) * exp(-1.96^2/2)});
	
	\node at (axis cs: 1.96, 0) [below] {$z_\alpha$};
	\draw[->] (axis cs: 2.2, 0.01) -- (axis cs: 2.5, 0.05) node[above] {$\alpha$};
	\end{axis}
\end{tikzpicture}
\end{center}

\subsubsection{Memoryless Property}
A random variable \( X \) is said to have the memoryless property if for all \( s, t \geq 0 \),
\[
	P(X > s + t | X > s) = P(X > t)
\]

Typical examples of memoryless distributions are the geometric distribution and the exponential distribution.

\subsection{Binomial Distribution}
Binomial distribution is denoted by \( \Binomial(n, p) \), where \( n \) is the number of trials and \( p \) is the probability of success in each trial. The random variable \( X \) represents the number of successes in \( n \) independent Bernoulli trials.
PMF is given by
\[
	P(X=k) = \binom{n}{k} p^k (1-p)^{n-k}, \quad k=0,1,2,\ldots,n
\]
the expected value is $np$
\begin{proof}
	Let $Y_k$ be the indicator random variable for the $k$-th trial, which takes the
	value 1 if the $k$-th trial is a success and 0 otherwise. Then,
	 \[
		E[Y_k] = 1 \cdot P(Y_k=1) + 0 \cdot P(Y_k=0) = p
	\] 
	by the independence,
	\[
		E[X] = E\left[\sum_{k=1}^{n} Y_k\right] = \sum_{k=1}^{n} E[Y_k] = np
	\] 
\end{proof}
the variance is $np(1-p)$
\begin{proof}
	\[
		E[Y_k^2] = 1^2 \cdot P(Y_k=1) + 0^2 \cdot P(Y_k=0) = p
	\] 
	\[
		V[Y_k] = E[Y_k^2] - E[Y_k]^2 = p - p^2 = p(1-p)
	\] 
	by the independence,
	\[
		V[X] = V\left[\sum_{k=1}^{n} Y_k\right] = \sum_{k=1}^{n} V[Y_k] = np(1-p)
	\] 
\end{proof}

\subsection{Poisson Distribution}
Poisson distribution is denoted by \( \Poisson(\mu) \), or $P(\mu)$, where
 \( \mu \) is the average number of events in a fixed interval of time or space.
If the expected number of events in a given interval time length $t$ is $\lambda$
then $\mu=\lambda t$.
The random variable \( X \) represents the number of events occurring in that interval.
PMF is given by
\[
	P(X=k) = \frac{\mu^k e^{-\mu}}{k!}, \quad k=0,1,2,\ldots
\] 
The expected value and variance of Poisson distribution are both equal to \( \mu \)
\begin{proof}
\[
	E[X] = \mu e^{-\mu}\sum_{k=1}^{\infty} \frac{\mu^{k-1} }{(k-1)!} = 
	\mu e^{-\mu} e^{\mu} = \mu
\] 
\[
	E[X^2] = \mu e^{-\mu}\sum_{k=1}^{\infty} \frac{k-1+1}{(k-1)!} \mu^{k-1}
	= \mu e^{-\mu} \left( 
	\mu\sum_{k=2}^{\infty} \frac{\mu^{k-2}}{(k-2)!} + e^{\mu} \right)
	= \mu^2 + \mu
\] 
\[
	V[X] = E[X^2] - E[X]^2 = \mu
\] 
\end{proof}

\subsubsection{Relationship between Binomial and Poisson Distributions}
The Poisson distribution can be derived as a limiting case of the Binomial 
distribution when \( n \to \infty \) and \( p \to 0 \) such that \( np = \mu \) remains
constant. 
Meaning, we cut the time interval into \( n \) subintervals, and let \( p \) be the probability
of an event occurring in each subinterval, while keeping the expected number of events
$np$ equal to original average $\lambda t$.
In this case,
\[
	\begin{aligned}
		P(X=k) &= \lim_{n \to \infty} \frac{n!}{k!(n-k)!} \left(\frac{p}{1-p}\right)^k
		((1-p)^{1/p})^{np}\\
					 &= \lim_{n \to \infty} \frac{n(n-1)\cdots(n-k+1)}{k!} p^k e^{-np}\\
					 &= \lim_{n \to \infty} \frac{n^k}{k!} p^k e^{-\mu}
					 = \frac{\mu^k e^{-\mu}}{k!}
	\end{aligned}
\] 


\subsection{Hypergeometric Distribution}
The hypergeometric rv \( X \) represents the number of successes in \( n \) draws
without replacement from a finite population of size \( N \) that contains
exactly \( M \) successes. The PMF is given by
\[
	P(X=k) = \frac{\binom{M}{k}\binom{N-M}{n-k}}{\binom{N}{n}}, \quad k=0,1,2,\ldots,n
\] 
the expected value is \( n\frac{M}{N} \)
and the variance is
\[
	V[X] = \frac{N-n}{N-1} n \frac{M}{N} \left(1-\frac{M}{N}\right)
\] 
\begin{remark}
	If let \( p = \frac{M}{N} \) be the probability of success in a single draw, then
	the expected value is the same as the Binomial distribution, while the variance
	is smaller than the Binomial distribution by a factor of \( \frac{N-n}{N-1} \).
\end{remark}

\subsection{Negative Binomial Distribution}
The negative binomial rv \( X \) represents the number of trials needed to achieve
exactly \( r \) th successes in a sequence of independent Bernoulli trials with
probability of success \( p \). The PMF is given by
\[
	P(X=k) = \binom{k-1}{r-1} p^r (1-p)^{k-r}, \quad k=r,r+1,r+2,\ldots
\] 
The expected value is \( \frac{r}{p} \)
and the variance is \( \frac{r(1-p)}{p^2} \)

Especially, when \( r=1 \), the negative binomial distribution reduces to the geometric distribution.

\subsection{Normal Distribution}
Normal distribution is denoted by \( N(\mu, \sigma^2) \), where \( \mu \) is the mean and \( \sigma^2 \) is the variance.
The PDF is given by
\[
	f(x) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}},\quad -\infty < x < \infty
\]
\begin{proof}
	\[
		E[X] = \frac{1}{\sigma \sqrt{2\pi}} \int_{-\infty}^{\infty} (x-\mu) e^{-\frac{(x-\mu)^2}{2\sigma^2}} dx
		+ \frac{\mu}{\sigma \sqrt{2\pi}} \int_{-\infty}^{\infty} e^{-\frac{(x-\mu)^2}{2\sigma^2}} dx
		= 0 + \mu = \mu
	\] 
	\[
		V[X] = \frac{1}{\sigma \sqrt{2\pi}} \int_{-\infty}^{\infty} t^2 e^{-\frac{t^2}{2\sigma^2}} dt
		= \frac{1}{\sigma \sqrt{2\pi}} \int -\sigma^2 t \dd e^{-\frac{t^2}{2\sigma^2}} 
		= -\frac{\sigma}{\sqrt{2\pi}} \left[ t e^{-\frac{t^2}{2\sigma^2}} \right]_{-\infty}^{\infty} 
		+ \frac{\sigma}{\sqrt{2\pi}} \int_{-\infty}^{\infty} e^{-\frac{t^2}{2\sigma^2}} dt
		= 0 + \sigma^2 = \sigma^2
	\] 
\end{proof}

We define the standard normal distribution CDF as
\[
	\Phi(z) = \int_{-\infty}^{z} \frac{1}{\sqrt{2\pi}} e^{-\frac{t^2}{2}} dt
\] 
has the following properties:
\begin{itemize}
	\item $\Phi(-z) = 1 - \Phi(z)$ or equivalently, $\Phi(z) + \Phi(-z) = 1$. Besides,
		\[
			\Phi(a) + \Phi(b) 
			\begin{cases}
				=1 & a+b=0 \\
				>1 & a+b>0 \\
				<1 & a+b<0
			\end{cases}
		\] 
	\item $\Phi(0) = 0.5$
	\item $P(|X|\leq x) = 2\Phi(x) - 1$
		\begin{proof}
			\[
			P(-x\leq X \leq x) = \Phi(x) - \Phi(-x) = \Phi(x) - (1-\Phi(x)) = 2\Phi(x) - 1
			\] 
		\end{proof}
\end{itemize}

\subsubsection{Linearity of Normal Distribution}
If $X_1, X_2, \ldots, X_n$ are independent normal random variables (with possibly different means and variances), then
any linear combination of them is also normally distributed.

\subsection{Lognormal Distribution}
The lognormal distribution rv \( X \) is defined as \( X = e^Y \), where \( Y \) is
a normal distribution rv with mean \( \mu \) and variance \( \sigma^2 \). 

The expected value and variance of lognormal distribution are given by
\[
	E[X] = e^{\mu + \frac{\sigma^2}{2}}, \quad V[X] = (e^{\sigma^2} - 1) e^{2\mu + \sigma^2}
\] 
and CDF is given by
\[
	F_X(x) = P(X \leq x) = P(Y \leq \ln x) = \Phi\left(\frac{\ln x - \mu}{\sigma}\right)
\] 

\subsection{Uniform Distribution}
The uniform distribution rv \( X \) in the interval \( [A, B] \) 
has
\[
	E[X] = \frac{A+B}{2}, \quad V[X] = \frac{(B-A)^2}{12}
\] 
\subsection{Exponential Distribution}
The exponential distribution rv \( X \) with rate parameter \( \lambda > 0 \) has
PDF
\[
	f(x) = \lambda e^{-\lambda x}, \quad x \geq 0
\] 
with
\[
	E[X] = \frac{1}{\lambda}, \quad V[X] = \frac{1}{\lambda^2}
\] 
\begin{proof}
	\[
		E[X] = \int_{0}^{\infty} x \lambda e^{-\lambda x} dx
		= -x e^{-\lambda x} \Big|_{0}^{\infty} + \int_{0}^{\infty} e^{-\lambda x} dx
		= 0 + \frac{1}{\lambda} = \frac{1}{\lambda}
	\] 
	\[
		E[X^2] = \int_{0}^{\infty} x^2 \lambda e^{-\lambda x} dx
		= -x^2 e^{-\lambda x} \Big|_{0}^{\infty} + 2 \int_{0}^{\infty} x e^{-\lambda x} dx
		= 0 + 2 \cdot \frac{1}{\lambda^2} = \frac{2}{\lambda^2}
	\] 
	\[
		V[X] = E[X^2] - E[X]^2 = \frac{2}{\lambda^2} - \frac{1}{\lambda^2} = \frac{1}{\lambda^2}
	\] 
\end{proof}

\subsubsection{Relationship between Exponential and Poisson Distributions}
The exponential distribution can describe the time between events in a Poisson
process. Let $T$ denotes the time until the first event occurs(equivalently to the time between events
according to the time-homogeneous property of Poisson process), then the CDF of $T$ is
\[
	F_T(t) = 1-P(T > t) = 1-P(0 \text{ events occur in } [0, t]) = 1-P_{\Poisson}(X=0) = 1-e^{-\mu} 
	= 1-e^{-\lambda t}
\] 
Taking the derivative of the CDF, we get the PDF of $T$:
\[
	f_T(t) = \frac{d}{dt} F_T(t) = \lambda e^{-\lambda t}, \quad t \geq 0
\] 

\subsection{Chi-Square Distribution}
The chi-square distribution rv \( X \) with \( k \) degrees of freedom is defined
as the sum of the squares of \( k \) independent standard normal random variables:
\[
	X = \sum_{i=1}^{k} Z_i^2, \quad Z_i \sim N(0, 1)
\] 

\subsection{Superposition of Distributions Parameter}
If $X$ and $Y$ are two independent random variables with the same distribution, then

\begin{itemize}
	\item $X\sim \Binomial(m, p), Y\sim \Binomial(n, p) \implies X+Y \sim \Binomial(m+n, p)$
	\item $X\sim \Poisson(\mu_1), Y\sim \Poisson(\mu_2) \implies X+Y \sim \Poisson(\mu_1+\mu_2)$
	\item $X\sim N(\mu_1, \sigma_1^2), Y\sim N(\mu_2, \sigma_2^2) \implies X+Y \sim N(\mu_1+\mu_2, \sigma_1^2+\sigma_2^2)$
	\item $X\sim \ChiSquare(m), Y\sim \ChiSquare(n) \implies X+Y \sim \ChiSquare(m+n)$
	\item $X\sim \Exponential(\lambda_1), Y\sim \Exponential(\lambda_2) \implies \min\{X, Y\} \sim \Exponential(\lambda_1+\lambda_2)$
\end{itemize}

\begin{proof}
	Here we only prove the last one, the others is very easy to prove.
	\[
		F_{\min} = P(\min\{X, Y\} \leq t) 
		= 1 - P(X>t, Y>t) = 1 - P(X>t)P(Y>t) = 1 - e^{-(\lambda_1 + \lambda_2)t}
	\] 
	compare to the CDF of exponential distribution finish the proof.
\end{proof}

\section{Transformations of Random Variables}

Given a random variable \( X \) with PDF \( f_X(x) \), and a transformation
\(Y = g(X) \), and we want to find the PDF of \( Y \), denoted by \( f_Y(y) \).

A common method is to use CDF transformation. Let \( F_X(x) \) be the CDF of \( X \), then
the CDF of \( Y \) is given by solving the inequality
\[
	F_Y(y) = P(Y \leq y) = P(g(X) \leq y) 
\] 
and if \( g \) is a invertible and monotonic function, we can use so called
Jacobian/change of variable method. Let \( h = g^{-1} \) be the inverse function of \( g \), then
\[
	f_Y(y) = f_X(h(y)) \left| \frac{dh(y)}{dy} \right|
\] 

\begin{theorem}[Probability Integral Transformation]
	Given a continuous random variable \( X \) with CDF \( F(x) \), then the random variable
	\[
		Y = F(X)
	\] 
	has a uniform distribution on the interval \( [0, 1] \).
	\begin{proof}
		\[
			F_Y(y) = P(Y \leq y) = P(F(X) \leq y) = P(X \leq F^{-1}(y)) = F(F^{-1}(y)) = y
		\] 
		for $0<y<1$, hence $Y$ has a uniform distribution on the interval $[0, 1]$.
	\end{proof}
\end{theorem}
\begin{example}
	Let $X$ be a continuous random variable with PDF
	\[
	f(x) = \begin{cases}
		\frac{x}{2}, & 0<x<2 \\
		0, & \text{otherwise}
	\end{cases}
	\] 
	Try to find $P(F(X)>E(X)-1)$
	\begin{answer}
		Let $Y=F(X)$, then $Y$ has a uniform distribution on the interval $[0, 1]$.
		Since $E(X)=\frac{4}{3}$, we have
		\[
		P(F(X)>E(X)-1) = P(Y>1/3) = 1-P(Y\leq 1/3) = 1-1/3 = 2/3
		\]
	\end{answer}
	
\end{example}

\subsection{Special Results}
\subsubsection{Min/Max of Random Variables}

Let \( X_1, X_2, \ldots, X_n \) be independent random variables with CDFs \( F_{X_1}(x), 
F_{X_2}(x), \ldots, F_{X_n}(x) \), then

\begin{theorem}
	The CDF of maximum is given by
	\[
		F_{\max}(x) = P(\max\{X_1, X_2, \ldots, X_n\} \leq x) = \prod_{i=1}^{n} F_{X_i}(x)
	\] 
\end{theorem}
\begin{theorem}
	The CDF of minimum is given by
	\[
		F_{\min}(x) = P(\min\{X_1, X_2, \ldots, X_n\} \leq x) = 1 - \prod_{i=1}^{n} (1-F_{X_i}(x))
	\]
	\begin{proof}
		\[
		\begin{aligned}
			F_{\min}(x) &= P(\min\{X_1, X_2, \ldots, X_n\} \leq x) \\
			&= 1 - P(\min\{X_1, X_2, \ldots, X_n\} > x) \\
			&= 1 - P(X_1 > x, X_2 > x, \ldots, X_n > x) \\
			&= 1 - \prod_{i=1}^{n} P(X_i > x) 
		\end{aligned}
		\] 
	\end{proof}
\end{theorem}

Especially, if \( X_1, X_2, \ldots, X_n \) are independent and identically distributed (i.i.d.) random
variables with CDF \( F(x) \), then
\[
F_{\max}(x) = F(x)^n, \quad F_{\min}(x) = 1 - (1-F(x))^n
\] 
and the corresponding PDFs are given by
\[
	f_{\max}(x) = n F(x)^{n-1} f(x), \quad f_{\min}(x) = n (1-F(x))^{n-1} f(x)
\] 

\begin{theorem}
	Let $U=\max\{X, Y\}$ and $V=\min\{X, Y\}$, then
	\[
	U+V = X+Y, \quad UV = XY
	\] 
	and 
	\[
		E[U] + E[V] = E[X] + E[Y], \quad E[UV] = E[X]E[Y]
	\] 
\end{theorem}
\begin{example}
	Let $X$ and $Y$ be two independent random variables, $X\sim \Exponential(1 / 3), Y\sim \Exponential(1 / 6)$,
	find $E[\min\{X, Y\}]$ and $E[\max\{X, Y\}]$.
	\begin{answer}
		Using the property of Exponential distribution, we have
		\[
			U = \min\{X, Y\} \sim \Exponential(1/3 + 1/6) = \Exponential(1/2)
		\] 
		since
		\[
			E[U] + E[V] = E[X] + E[Y] = 3 + 6 = 9
		\] 
		we have
		\[
			E[V] = 9 - E[U] = 9 - 2 = 7
		\] 
	\end{answer}
\end{example}

\subsubsection{Absolute Value of Normal Distribution}
\begin{theorem} Let \( X \sim N(0, \sigma^2) \), then
	\[
		E(|X|) = \sqrt{\frac{2}{\pi}}\sigma, \quad V(|X|) = \left(1 - \frac{2}{\pi}\right)\sigma^2
	\] 
\end{theorem}

\subsubsection{Transformation from Min/Max to Absolute Value}
\[
	\max\{X, Y\} = \frac{X+Y+|X-Y|}{2}, \quad \min\{X, Y\} = \frac{X+Y-|X-Y|}{2}
\] 
\subsubsection{Permutation symmetry}
If \( X_1, X_2, \ldots, X_n \) are i.i.d. random variables, then 
for some expressions we can use permutation symmetry.
\begin{theorem}
	\[
		E\left[\frac{X_i^2}{\sum_{j=1}^{n} X_j^2}\right] = \frac{1}{n}
	\] 
	Especially, when \( n=2 \), we have
\[
		E\left[\frac{X_1^2}{X_1^2+X_2^2}\right] = E\left[\frac{X_2^2}{X_1^2+X_2^2}\right] = \frac{1}{2}
\] 
\end{theorem}

\section{Covariance and Variance}

\begin{theorem}
	\[
	\Cov(X+c, Y) = \Cov(X, Y)
	\] 
	\begin{proof}
		\[
			\Cov(X+c, Y) = E[(X+c)Y] - E[X+c]E[Y] = E[XY]+cE[Y] - (E[X]+c)E[Y] = \Cov(X, Y)
		\]
	\end{proof}
\end{theorem}

\begin{theorem}
	\[ \Cov(aX+bY, Z) = a\Cov(X, Z) + b\Cov(Y, Z) \] 
	\begin{proof}
		\[
			\begin{aligned}
				\Cov(aX+bY, Z) &= E[(aX+bY)Z] - E[aX+bY]E[Z] \\
				&= aE[XZ] + bE[YZ] - (aE[X]+bE[Y])E[Z] \\
				&= a(E[XZ]-E[X]E[Z]) + b(E[YZ]-E[Y]E[Z]) \\
				&= a\Cov(X, Z) + b\Cov(Y, Z)
				
			\end{aligned}
		\]
	\end{proof}
\end{theorem}

\begin{theorem}
	\[
		V(X\pm Y) = V(X) + V(Y) \pm 2\Cov(X, Y)
	\] 
	or equivalently,
	\[
		\sigma^2_{X\pm Y} = \sigma^2_X + \sigma^2_Y \pm 2\rho\sigma_X\sigma_Y
	\] 
\end{theorem}

\begin{theorem}
	If \( X \) and \( Y \) are independent, then
	\[
		V(XY) = V(X)V(Y) + V(X)E[Y]^2 + V(Y)E[X]^2
	\] 
		\begin{proof}
		\[
			\begin{aligned}
				V(XY) &= E[X^2Y^2] - E[XY]^2 \\
				&= E[X^2]E[Y^2] - E[X]^2E[Y]^2 \\
				&= (V(X)+E[X]^2)(V(Y)+E[Y]^2) - E[X]^2E[Y]^2
			\end{aligned}
		\]
		\end{proof}
\end{theorem}
\begin{remark}
	Actually, we seldom recite this formula, but remember how to derive it, and
	use the derivation for each problem.
\end{remark}

\begin{theorem}
	\[
	E(XY) = E(X)E(Y) \iff \Cov(X, Y) = 0
	\] 
\end{theorem}

Sometimes, when not independent, we can try to decompose the random variables into
independent parts.
\begin{example}
	Let r.v. $X_1, X_2, \ldots, X_n$ be i.i.d. with variance $\sigma^2$, then let $Y = \frac{1}{n}\sum_{i=1}^{n} X_i$, 
we have
\[
	V(X_1+ Y) = V(X_1 + 1/n \sum_{i=1}^{n} X_i) = V\left((1+\frac{1}{n})X_1 + \frac{1}{n} \sum_{i=2}^{n} X_i\right)
	= \left(\frac{n+1}{n}\right)^2 \sigma^2 + \frac{n-1}{n^2} \sigma^2 
	= \frac{n+3}{n} \sigma^2
\] 
	
\end{example}

\section{2D Distributions}

\subsection{Marginal Distribution}

We define the marginal distributions of \( X \) and \( Y \) as
\[
	f_X(x) = \int_{-\infty}^{\infty} f_{X, Y}(x, y) dy, \quad
	f_Y(y) = \int_{-\infty}^{\infty} f_{X, Y}(x, y) dx
\] 
If given a 2D CDF \( F(x, y) \), then the marginal CDFs are given by
\[
	F_X(x) = \lim_{y \to \infty} F(x, y), \quad
	F_Y(y) = \lim_{x \to \infty} F(x, y)
\] 

\begin{example}
	The $F_Y(y)$ for given 2D CDF
	 \[
		 F(x, y) = \frac{1}{2} \left(\Phi(x)\Phi(y) + \Phi(\min\{x, y\})\right)
	\] 
	is given by
	\[
		F_Y(y) = \lim_{x \to \infty} F(x, y) = \frac{1}{2} \left(\Phi(y) + \Phi(y)\right) = \Phi(y)
	\]
	
\end{example}

\subsection{2D Normal Distribution}
The 2D normal distribution PDF is given by
\[
	f(x, y) = \frac{1}{2\pi \sigma_x \sigma_y \sqrt{1-\rho^2}}
	\exp\left\{-\frac{1}{2(1-\rho^2)} \left[ \left(\frac{x-\mu_x}{\sigma_x}\right)^2
	- 2\rho \left(\frac{x-\mu_x}{\sigma_x}\right)\left(\frac{y-\mu_y}{\sigma_y}\right) + \left(\frac{y-\mu_y}{\sigma_y}\right)^2 \right] \right\}
\] 
where $|\rho|<1$ denotes the correlation coefficient between $X$ and $Y$,
\[
	\rho = \frac{\Cov(X, Y)}{\sigma_x \sigma_y}
\] 
and we call denote
\[
	(X, Y) \sim N(\mu_x, \mu_y; \sigma_x^2, \sigma_y^2; \rho)
\] 

\begin{theorem}
	When given a 2D distribution
	\[
		f(x, y) = \frac{1}{\text{Constant}} \exp\left\{Ax^2 + Bxy + Cy^2 + Dx + Ey + F\right\}
	\] 
	when it is a 2D normal distribution, then
	\[
		\rho^2 = \frac{B^2}{4AC}
	\] 
	while the sign of $\rho$ is the same as the sign of $B$.
	\begin{proof}
		Let $G=-1 / (2(1-\rho^2))$, then we have
		\[
			A = \frac{G}{\sigma_x^2}, \quad B = -\frac{2G\rho}{\sigma_x\sigma_y}, \quad C = \frac{G}{\sigma_y^2}
		\] 
	\end{proof}
\end{theorem}
\begin{remark}
	After solving the $\rho$, we can subtitute it back to get  $G$ and then $\sigma_x$ and $\sigma_y$.
\end{remark}

\begin{example}
	For 2D normal distribution
	\[
		f(x, y) = \frac{1}{2\pi} \exp\left\{ -x^2 + xy - \frac{y^2}{2} \right\}
	\] 
	we have
	\[
		\rho^2 = \frac{1}{2} \implies \rho = \frac{\sqrt{2}}{2}
	\] 
	hence
	\[
		G = -\frac{1}{2(1-\rho^2)} = -1
	\] 
	then we can extract $G$ from exponent
	\[
		f(x, y) = \frac{1}{2\pi} \exp\left\{ -1 \left[ x^2 - \sqrt{2}xy + \frac{y^2}{2} \right] \right\}
	\] 
	or directly use the G formula to get
	\[
		\sigma_x^2 = \frac{1}{-G} = 1, \quad \sigma_y^2 = \frac{1}{-2G} = 2
	\] 
\end{example}

\subsubsection{Properties of 2D Normal Distribution}

\begin{theorem}
	For 2D normal distribution, 
	\[
		\rho = 0 \iff X \text{ and } Y \text{ are independent}
	\] 
\end{theorem}
\begin{theorem}
Two marginal distributions of 2D normal distribution are also normal distributions,
while the inverse is not true(only true when $\rho=0$, i.e. $X$ and $Y$ are independent).
\begin{proof}
	A simple counterexample is $X\sim N(0, 1), Y=2X$, then $X$ and $Y$ are both
	normal distributions, but $(X, Y)$ is not a 2D normal distribution.
\end{proof}
\end{theorem}

In general, the linear combination of two normal distributions is not necessarily a normal distribution.
\begin{theorem}
	The linear combination of two marginal distributions of 2D normal distribution 
is also a normal distribution.
\[
aX+bY \sim N(a\mu_x+b\mu_y, a^2\sigma_x^2 + b^2\sigma_y^2 + 2ab\rho\sigma_x\sigma_y)
\] 
\end{theorem}
Especially, this theorem implies that if $X$ and $Y$ are independent, then the
linear combination of them is also a normal distribution.

\begin{example}
	Let $(X, Y) \sim N(0, 0; 1, 4; -1/2)$
	then
	\[
	X+Y \sim N(0, 1 + 4 + 2\cdot(-1/2)\cdot 1 \cdot 2) = N(0, 3)
	\] 
	\[
	X-Y \sim N(0, 1 + 4 - 2\cdot(-1/2)\cdot 1 \cdot 2) = N(0, 7)
	\] 
	
\end{example}

\begin{theorem}
	After invertible linear transformation, the 2D normal distribution is still a 2D normal distribution.
\end{theorem}

\begin{example}
	Let $(X, Y)$ obeys a 2D normal distribution, try to show that $Z = X+Y$ also obeys a normal distribution.
	\begin{answer}
		\[
		\begin{pmatrix}
			X+Y \\ Y
		\end{pmatrix} =
		\begin{pmatrix}
			1 & 1 \\ 0 & 1
		\end{pmatrix}
		\begin{pmatrix}
			X \\ Y
		\end{pmatrix}
		\] 
		since the transformation matrix is invertible, we have $Z = X+Y$ also obeys a normal distribution.
	\end{answer}
\end{example}

\section{Law of Large Numbers and Central Limit Theorem}

\subsection{Markov's Inequality and Chebyshev's Inequality}
\begin{lemma}[Markov's Inequality]
	Let $X$ be a \textbf{non-negative} (i.e $X \geq 0$) random variable with finite expected value $\mu$, then
	for any real number $a > 0$ we have
	\[
		P(X \geq a) \leq \frac{\mu}{a}
	\]
	\begin{proof}
		Here we give a not very rigorous proof(not considering mixed discrete and continuous and other edge cases).
		\[
			a P(X \geq a) = \int_a^{\infty} a f_X(x) dx \leq \int_a^{\infty} x f_X(x) dx
			\leq \int_0^{\infty} x f_X(x) dx = E[X] = \mu
		\]
		dividing both sides by $a$ gives the result.
	\end{proof}
\end{lemma}

\begin{theorem}[Chebyshev's Inequality]
	Let $X$ be a random variable with finite expected value $\mu$ and finite non-zero variance $\sigma^2$, then
	for any real number $\epsilon$ we have
	 \[
		 P\{|X-\mu| \geq \epsilon\} \leq \frac{\sigma^2}{\epsilon^2}
	\] 
	or equivalently,
	\[
		P\{|X-\mu| < \epsilon\} > 1 - \frac{\sigma^2}{\epsilon^2}
	\] 
\end{theorem}
\begin{proof}
	Let $Y = (X-\mu)^2$, then $Y$ is a non-negative random variable with expected value $\sigma^2$.
	Applying Markov's inequality to $Y$, we have
	\[
		P(Y \geq \epsilon^2) \leq \frac{\sigma^2}{\epsilon^2}
	\] 
	which is equivalent to
	\[
		P(|X-\mu| \geq \epsilon) \leq \frac{\sigma^2}{\epsilon^2}
	\]
\end{proof}
\begin{remark}
	Another useful form of Chebyshev's inequality is to let $\epsilon = k\sigma$, then we have
	\[
		P(|X-\mu| \geq k\sigma) \leq \frac{1}{k^2}
	\] 
\end{remark}

\begin{example}
	Known that $\mu_X=-2, \mu_2, \sigma^2_X=1, \sigma^2_Y=4$ and the correlation
	coefficient $\rho = -1/2$, try to find the upper bound of $P(|X+Y| \geq 6)$.
	\begin{answer}
		\[
			E[X+Y] = \mu_X + \mu_Y = -2 + 2 = 0
		\]
		\[
			V[X+Y] = V[X] + V[Y] + 2\rho\sigma_X\sigma_Y = 1 + 4 + 2\cdot(-1/2)\cdot 1 \cdot 2 = 3
		\]
		\[
			P(|X+Y| \geq 6) \leq \frac{V[X+Y]}{6^2} = \frac{3}{36} = \frac{1}{12}
		\] 
\end{example}

\subsection{Law of Large Numbers}
\begin{theorem}[Chebyshev's Law of Large Numbers]
	If $X_1, X_2, \ldots, X_n$ are independent random variables with finite expected value $\mu_i$ and
	finite variance $\sigma_i^2$, then for any $\epsilon > 0$ we have
	\[
		\lim_{n \to \infty} P\left\{\left|\frac{1}{n}\sum_{i=1}^{n} X_i - \frac{1}{n}\sum_{i=1}^{n} \mu_i\right| < \epsilon\right\} = 1
	\]
	especially, if $X_1, X_2, \ldots, X_n$ has same expected value $\mu$ we have
	\[
		\lim_{n \to \infty} P\left\{\left|\frac{1}{n}\sum_{i=1}^{n} X_i - \mu\right| < \epsilon\right\} = 1
	\]
	\begin{remark}
		Chebyshev's Law of Large Numbers does not require the random variables to be
		identically distributed.
	\end{remark}
	\begin{remark}
		In a not very rigorous sense, we can say that the as $n \to \infty$, the sample
		mean will converge to the expected value in probability. The word
	  ``in probability" means that we can not guarantee that the sample mean actually
		converges to the expected value, but we can guarantee that the probability of
		the sample mean does not converge to the expected value is zero.
	\end{remark}
\end{theorem}

\subsection{Central Limit Theorem}
\begin{theorem}[Levy-Lindeberg Central Limit Theorem]
	If $X_1, X_2, \ldots, X_n$ are i.i.d. random variables with finite expected
	value $\mu$ and finite variance $\sigma^2$, then for any real number $x$ we have
	\[
		\lim_{n \to \infty} P\left\{\frac{\sum_{i=1}^{n} X_i - n\mu}{\sigma\sqrt{n}} \leq x\right\} = \Phi(x)
	\] 
	i.e. the distribution of the standardized sum of i.i.d. r.v. converges to the standard normal distribution as $n \to \infty$.
	In other words, the sum of i.i.d. r.v. converges to the normal distribution as $n \to \infty$.
\end{theorem}

\section{Integral Results}
\subsection{Poisson Integral}
\[
	\int_{-\infty}^{\infty} e^{-x^2} dx = \sqrt{\pi}
\] 
\[
	\int_{-\infty}^{\infty} e^{-x^2/2} dx = \sqrt{2\pi}
\] 
\subsection{Exponential Integral}
\[
	\int_{0}^{\infty} e^{-\alpha x} dx = \frac{1}{\alpha}, \quad \alpha > 0
\] 

\subsection{Gamma Function}
%TODO

\end{document}
